### 2013-07-24 doc/asplos14/evaluation.tex
Incorporate new UTS plot, add todo to rewrite text to match
eval update

The goal of our evaluation is to show that the core pieces of the \Grappa
runtime system, namely our tasking system and the global memory/communication
layer work as expected and together are able to efficiently run irregular
applications. We evaluate \Grappa in three basic steps:

@@

\item We present results that show that \Grappa can
    support large amounts of concurrency, sufficient for remote memory
    access and aggregation. The communication layer is able to
    sustain a very high rate of global memory operations. We also show the performance of a
graph kernel that stresses communication and concurrency together.
\TODO{compare to key-value stores?}

\item We characterize system behavior, including
profiling where execution time goes, how aggregation affects message size and
rates, and \checkme{the relationship between aggregation and
    concurrency.}

@@

\TODO{Grappa is well-suited to applications where there is
    abundant data parallelism but unpredictable or imbalanced data
    access. BSP is efficient where there is sufficient parallel slack.
    Although Grappa loses some asynchrony by relying heavily on
    aggregation, it still provides more flexibility for sections of the
    computation to proceed more quickly. }

\TODO{how much concurrency is needed on apps. Our basic plot of
    performance vs workers (with and without aggro?)}
\TODO{relationship between user-thread-visible latency and aggregate throughput.}
\TODO{characterization of aggregation message size vs nodes}
\TODO{show our scalability plots; if they don't scale, show
    specifically why. It is okay if its because aggregation stops
    working after a point if we can discuss future scalability.
    Perhaps reference ``MPI on millions of cores'' type work}

\end{itemize}

@@

If we calculate aggregate context switch \emph{rate\/} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4~cache lines (1~for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
\TODO{based on this and GUPS, our message rate will be limited by
    context switching for small+single messages. One experiment is
    GUPs with context switching, where we vary number of reads/suspend}

In summary, our context switch engine is able to efficiently sustain very high
concurrency and, as we will show later, the amount of concurrency sustained is
sufficient for the latencies \Grappa needs to hide.

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark, which
measures cluster-wide random access bandwidth.  Read-modify-write
updates are dispatched at random to a large array. This benchmark
stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain
well over a billion updates per second with 64~nodes. Note also that
when aggregation is turned off, the update rate is nearly
flat. Clearly aggregation is instrumental for good communication
performance.

@@

\TODO{go into more detail about what Grappa is doing when executing
    GUPs. Specifically note what the rate per core is, what the
    aggregation and deaggregation rates are. Show on the plot the
    message size distribution (some box plot)}

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput -- at the 16-node data point,
the average active tasks per core over the search is 775 for T1 and
only 13 for T3.
On the other hand, \Grappa performs and scales very well for T1
trees.

@@

\paragraph{BFS.}
The Grappa and MPI BFS implementations are quite similar. They both
use the same graph representation, and they both depend on aggregation
for performance. The difference is that the BFS' aggregation is
integrated with the traversal code and specialized for the problem:
since the only messages required are edge traversals, the aggregation
buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
pays execution and space overhead in order to support aggregation of
arbitrary closures; the messages \Grappa aggregates include the two
8-byte vertex IDs along with 8 bytes of deserialization information as
well as 8 bytes of synchronization information. Furthermore, the MPI
implementation writes the edge information directly into the
aggregation buffer while \Grappa executes additional code to serialize
and deserialize the messages. \TODO{Include message counts, size, and
  rates; code size too?}.

@@

\paragraph{What limits \Grappa's performance?}
The most common operation in \Grappa is sending a message. Three key
operations occur in a message's lifetime: creating and enqueuing a
message, serializing a message in the aggregator, and deserializing a
message at the destination. We benchmarked each of these steps
individually to find the bottleneck.

Minimum-sized messages can be created and enqueued at a rate of 16M/s,
serialized into an aggregation buffer at 32M/s, and deserialized from
an aggregation buffer at 210M/s.  Together, these rates limit us to
$~$10M/s per core in the best case. In a blocking application, the
context switch costs (50ns, or 20M/s) bring this maximum down to
7M/s. In practice, \Grappa does not achieve this maximum; the current
aggregator design communicates inefficiently between cores within a
node, limiting our message rate to to $~$1M/s per core with messages
of common sizes. Future work will reduce this cost.

>>>>

The goal of our evaluation is to understand whether the core pieces of the \Grappa
runtime system, namely our tasking system and the global memory/communication
layer work as expected and whether together they are able to efficiently run irregular
applications. We evaluate \Grappa in three basic steps:

@@

\item We present results that show that \Grappa can
    support large amounts of concurrency, sufficient for remote memory
    access and aggregation. The communication layer is able to
    sustain a very high rate of global memory operations. We also show the performance of a
graph kernel that stresses communication and concurrency together.
%\TODO{compare to key-value stores?}

\item We characterize system behavior, including
profiling where execution time goes, and how aggregation affects message size and
rates.

%, and \checkme{the relationship between aggregation and
%    concurrency.}

@@

\comment{
\TODO{Grappa is well-suited to applications where there is
    abundant data parallelism but unpredictable or imbalanced data
    access. BSP is efficient where there is sufficient parallel slack.
    Although Grappa loses some asynchrony by relying heavily on
    aggregation, it still provides more flexibility for sections of the
    computation to proceed more quickly. }

\TODO{how much concurrency is needed on apps. Our basic plot of
    performance vs workers (with and without aggro?)}
\TODO{relationship between user-thread-visible latency and aggregate throughput.}
\TODO{characterization of aggregation message size vs nodes}
\TODO{show our scalability plots; if they don't scale, show
    specifically why. It is okay if its because aggregation stops
    working after a point if we can discuss future scalability.
    Perhaps reference ``MPI on millions of cores'' type work}
}
\end{itemize}

@@

If we calculate aggregate context switch \emph{rate\/} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4~cache lines (1~for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).
%\TODO{based on this and GUPS, our message rate will be limited by
%    context switching for small+single messages. One experiment is
%    GUPs with context switching, where we vary number of reads/suspend}

In summary, our tasking layer is able to efficiently sustain very high
concurrency and, as we will show later, the amount of concurrency sustained is
sufficient for the latencies \Grappa needs to hide.

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark, which
measures cluster-wide random access bandwidth.  Read-modify-write
updates are dispatched at random to a global large array. This benchmark
stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain
well over a billion updates per second with 64~nodes. Note also that
when aggregation is turned off, the update rate is nearly
flat. Clearly aggregation is instrumental for good communication
performance.

@@

%\TODO{go into more detail about what Grappa is doing when executing
%    GUPs. Specifically note what the rate per core is, what the
%    aggregation and deaggregation rates are. Show on the plot the
%    message size distribution (some box plot)}

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \comment{\TODO{cite Valiant and say more somewhere about BSP}}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput -- at the 16-node data point,
the average active tasks per core over the search is 775 for T1 and
only 13 for T3.
On the other hand, \Grappa performs and scales very well for T1
trees.

@@

\paragraph{BFS.}
The Grappa and MPI BFS implementations are quite similar. They both
use the same graph representation, and they both depend on aggregation
for performance. The difference is that the BFS' aggregation is
integrated with the traversal code and specialized for the problem:
since the only messages required are edge traversals, the aggregation
buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
pays execution and space overhead in order to support aggregation of
arbitrary messages; the messages \Grappa aggregates include the two
8-byte vertex IDs along with 8 bytes of deserialization information as
well as 8 bytes of synchronization information. Furthermore, the MPI
implementation writes the edge information directly into the
aggregation buffer while \Grappa executes additional code to serialize
and deserialize the messages.

@@

\paragraph{What limits \Grappa's performance?}
The most common operation in \Grappa is sending a message. Three key
operations occur in a message's lifetime: creating and enqueuing a
message, serializing a message in the aggregator, and deserializing a
message at the destination. We benchmarked each of these steps
individually to shed light on the mechanism.

Minimum-sized messages can be created and enqueued at a rate of 16M/s,
serialized into an aggregation buffer at 32M/s, and deserialized from
an aggregation buffer at 210M/s.  Together, these rates limit us to
$~$10M/s per core in the best case. In a blocking application, the
context switch costs (50ns, or 20M/s) bring this maximum down to
7M/s. In practice, \Grappa does not achieve this maximum; the current
aggregator design communicates inefficiently between cores within a
node, limiting our message rate to to $~$1M/s per core with messages
of common sizes. Future work will reduce this cost.  \comment{Ack! -Mark}
